On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-04-28 Manuscript/manuscript.tex
minor comments

%\jy{Avoid references in abstract.}

\jy{This is only the background. An abstract also needs to summarize the 
contributions.}
%\bigskip
%\eds{removed Devon Allen from keywords because his name is already in title
%and replaced with others}

@@

\noindent\sc{Keywords}:
False Start, Hurdles, Reaction Time, Seiko

@@

\jy{Turn on column numbers in your editor to keep line width under 80.}

@@

\jy{Not a good opening. Could start by telling the story, recapping the major
  events and discussions on the internet. Then dig into the relevant
  literature. Finally outline the contributions of this work.
}

\jy{use the following structure and flow:
  1) A paragraph to tell the story;
  2) A paragraph on informal statistical analyses on the internet/social media;
  3) A paragraph on scholarly articles on reaction time in the literature;
  4) A paragraph on what's new: methods; results.
  5) A paragraph of roadmap.}

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

@@

\jy{Could be condensed}
\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and at his physical peak.

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
%\eds{are you using covariates?  Data section
%implies that you are, but Results section does not}
%\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

@@

\jy{You have two such datasets in the analysis. Use one paragraph for each.}
\of{Split it into 2 sections, do you still want me to do something else?}

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
how is this athlete clustering being accounted for here?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
Gamma family with an identity link works well for their reaction-time data.
It doesn't seem like a good idea in general because it is likely that some 
negative predicted responses will occur either during model fitting or in the 
final model, but the internet seems to favor this link so a reviewer may ask 
about it too.}
We include a random effect for venue because the technology
that World Athletics has used has differed since 1999, and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Additionally, any other weather or climate related factors 
such as humidity, precipitation, elevation, could be captured by this effect.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

>>>>

%\eds{removed Devon Allen from keywords because his name is already in title
%and replaced with others}

@@

\bigskip\noindent\sc{Keywords}:
False Start, Hurdles, Reaction Time, Seiko

@@

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

@@

\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and at his physical peak.

@@

%4)
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
%\eds{are you using covariates?  Data section
%implies that you are, but Results section does not}
%\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

@@

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
  how is this athlete clustering being accounted for here?}
\jy{We could add an athlete random effect too, if the players are identified in the data.}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
Gamma family with an identity link works well for their reaction-time data.
It doesn't seem like a good idea in general because it is likely that some 
negative predicted responses will occur either during model fitting or in the 
final model, but the internet seems to favor this link so a reviewer may ask 
about it too.}
We include a random effect for venue because the technology
that World Athletics has used has differed since 1999, and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Additionally, any other weather or climate related factors 
such as humidity, precipitation, elevation, could be captured by this effect.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.
